---
title: "Screening peer-reviewer candidates with coiCheckR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Screening peer-reviewer candidates with coiCheckR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("COICHECKR_RUN_LIVE_TESTS"), "true")
)
```

## Motivation

Editors screening peer-reviewer candidates usually check for direct
co-authorship with a manuscript's authors and stop there. In a small
field, that misses two common conflict patterns: a candidate who
shares a *frequent collaborator* with an author (second-degree
conflict), and a candidate who shares *federal funding* with an author
with no joint publication yet. `coiCheckR` checks all three and
returns evidence for each -- not a conflicted/clear verdict, since
that judgment call should stay with the editor.

This vignette uses a real, confirmed conflict as its running example:
the package author, Janani Ravi (`"Ravi J"`), and her CU Anschutz
colleague Arjun Krishnan (`"Krishnan A"`), who have co-published and
hold NIH funding together. A screen against this pair is expected to
flag on all three evidence types -- direct co-authorship, second-degree
overlap, and shared NIH funding. A fictitious candidate, `"Doe C"`, is
included alongside as a clean comparison.

```{r setup}
library(coiCheckR)
```

## A single candidate, all three checks

```{r single-candidate}
report <- checkCoi(
  candidate_name = "Ravi J",
  author_names = "Krishnan A",
  affiliation = "Colorado",
  check_second_degree = TRUE,
  check_funding = TRUE
)
report
```

`report$direct` holds one row per shared publication (PMID and year),
`report$second_degree` one row per collaborator-mediated path, and
`report$funding` one row per shared NIH award -- each independently
verifiable evidence, not a verdict.

## Screening a shortlist at once

```{r batch}
batch <- checkCoiBatch(
  candidate_names = c("Ravi J", "Doe C"),
  author_names = "Krishnan A",
  affiliation = "Colorado",
  check_second_degree = TRUE,
  check_funding = TRUE
)

batch$summary
```

`formatCoiSummary()` turns the same result into plain-text lines
suitable for pasting into an email to a co-editor:

```{r format-summary}
cat(formatCoiSummary(batch), sep = "\n")
```

## Shared NIH funding

Some conflicts show up in funding before they show up in the
literature -- e.g. a shared multi-PI award with no joint publication
yet. `reporterSharedAwards()` cross-references NIH RePORTER project
numbers directly:

```{r funding}
this_year <- as.integer(format(Sys.Date(), "%Y"))
last_10y <- (this_year - 9):this_year

reporterSharedAwards("Ravi, Janani", "Krishnan, Arjun", fiscal_years = last_10y)
```

## Visualizing a batch result

```{r plot, fig.width = 6, fig.height = 6}
reports <- batch[setdiff(names(batch), "summary")]
plotCoiNetwork(reports)
```

## A note on evidence vs. verdicts

Every function in this package returns evidence -- PMIDs, award
numbers, collaborator names -- rather than a binary conflicted/clear
flag. Two candidates with the same number of `direct` hits can warrant
different editorial decisions depending on recency, the nature of the
collaboration, and context the tool has no way to know. `coiCheckR` is
meant to make that judgment call easier to make well, not to make it
for you.
