---
title: "Screening peer-reviewer candidates with coiCheckR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Screening peer-reviewer candidates with coiCheckR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("COICHECKR_RUN_LIVE_TESTS"), "true")
)
```

## Motivation

Editors screening peer-reviewer candidates usually check for direct
co-authorship with a manuscript's authors and stop there. In a small
field, that misses two common conflict patterns: a candidate who
shares a *frequent collaborator* with an author (second-degree
conflict), and a candidate who shares *federal funding* with an author
with no joint publication yet. `coiCheckR` checks all three and
returns evidence for each -- not a conflicted/clear verdict, since
that judgment call should stay with the editor.

This vignette uses illustrative placeholder names (`"Smith AB"`,
`"Lee C"`, `"Doe C"`) as its running example -- substitute a candidate
and author pair you know to share co-authorship and NIH funding to see
a live screen flag on all three evidence types: direct co-authorship,
second-degree overlap, and shared NIH funding. `"Doe C"` is included
alongside as a clean comparison with no expected hits.

Live network calls in this vignette are gated behind the
`COICHECKR_RUN_LIVE_TESTS` environment variable and do not run during
package build/check, so the placeholder names below produce no output
unless you build this vignette locally with real names and that
variable set.

```{r setup}
library(coiCheckR)
```

## A single candidate, all three checks

```{r single-candidate}
report <- checkCoi(
  candidate_name = "Smith AB",
  author_names = "Lee C",
  check_second_degree = TRUE,
  check_funding = TRUE
)
report
```

`report$direct` holds one row per shared publication (PMID and year),
`report$second_degree` one row per collaborator-mediated path, and
`report$funding` one row per shared NIH award -- each independently
verifiable evidence, not a verdict.

## Screening a shortlist at once

```{r batch}
batch <- checkCoiBatch(
  candidate_names = c("Smith AB", "Doe C"),
  author_names = "Lee C",
  check_second_degree = TRUE,
  check_funding = TRUE
)

batch$summary
```

`formatCoiSummary()` turns the same result into plain-text lines
suitable for pasting into an email to a co-editor:

```{r format-summary}
cat(formatCoiSummary(batch), sep = "\n")
```

## Shared NIH funding

Some conflicts show up in funding before they show up in the
literature -- e.g. a shared multi-PI award with no joint publication
yet. `reporterSharedAwards()` cross-references NIH RePORTER project
numbers directly:

```{r funding}
this_year <- as.integer(format(Sys.Date(), "%Y"))
last_10y <- (this_year - 9):this_year

reporterSharedAwards("Smith, Anne", "Lee, Charles", fiscal_years = last_10y)
```

## Visualizing a batch result

```{r plot, fig.width = 6, fig.height = 6}
reports <- batch[setdiff(names(batch), "summary")]
plotCoiNetwork(reports)
```

## A note on evidence vs. verdicts

Every function in this package returns evidence -- PMIDs, award
numbers, collaborator names -- rather than a binary conflicted/clear
flag. Two candidates with the same number of `direct` hits can warrant
different editorial decisions depending on recency, the nature of the
collaboration, and context the tool has no way to know. `coiCheckR` is
meant to make that judgment call easier to make well, not to make it
for you.

## Session info

```{r session-info}
sessionInfo()
```
