---
title: "A function-by-function walkthrough of coiCheckR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A function-by-function walkthrough of coiCheckR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("COICHECKR_RUN_LIVE_TESTS"), "true")
)
```

## Why this vignette exists

`vignette("coiCheckR")` shows the package the way an editor would use
it -- start with `checkCoi()`, get a report. This vignette instead
walks through the pieces `checkCoi()` is built from, one at a time, in
the order they naturally compose: raw PubMed search, then fetching
metadata for those results, then building a co-authorship graph from
that metadata, then the higher-level conflict checks, then the
composite report, then presentation/visualization. It's meant for
verifying the package's plumbing -- if something looks wrong in a
`checkCoi()` result, this walkthrough helps narrow down *which layer*
produced it.

Live network calls here are gated behind the `COICHECKR_RUN_LIVE_TESTS`
environment variable and do not run during package build/check, same
as `vignette("coiCheckR")`. Set it (plus real names below) to run this
locally.

```{r setup}
# If you have this repo cloned locally (e.g. you're reviewing a PR
# branch), load it directly -- this reflects exactly what's on disk, no
# network involved. Note the path: knitr evaluates vignette chunks with
# the *vignette's own directory* as the working directory, so this is
# ".." (the package root), not ".":
devtools::load_all("..")

# Otherwise, to install from GitHub -- devtools::install_github() is
# still the current, supported way to do this (it delegates to
# remotes::install_github() under the hood). Installing plain
# "JRaviLab/coiCheckR" gets the *default* branch (main); pass `ref` to
# install a specific PR branch instead:
# devtools::install_github(
#   "JRaviLab/coiCheckR",
#   ref = "fix/bioconductor-packaging-and-live-api-bugs"
# )
# library(coiCheckR)
```

```{r names}
# Optional: raise the PubMed rate limit from 3 to 10 req/sec if you're
# going to run more than a handful of these (free key, no account
# needed beyond an NCBI login):
# https://www.ncbi.nlm.nih.gov/account/settings/
# setNCBIKey("your-ncbi-api-key")

candidate <- "Ravi Janani" # the reviewer candidate being screened. A full
                            # forename here is fine -- and searches PubMed
                            # *more* precisely than bare initials (34 hits
                            # vs. 105 for "Ravi J") -- because the package
                            # internally reduces every name to "Last F" for
                            # matching (see .pm_canonical_name(), and step 4
                            # below for exactly where that happens)

# 34 hits for "Ravi Janani" is still not unique enough on its own.
# `affiliation` narrows the underlying PubMed search to a substring
# match on the recorded affiliation string -- and since that reflects
# whatever institution was listed *at the time of that publication*,
# not a current affiliation, someone with a multi-institution career
# needs every one listed, one per vector element (23/34 papers, filtered
# this way), or earlier-career papers are silently dropped.
#
# Two DIFFERENT things are going on here, worth not conflating:
#   - ACROSS vector elements, this is OR: any paper matching Colorado,
#     OR Michigan, OR Rutgers, OR Virginia is included.
#   - WITHIN one element, PubMed matches it as a literal phrase, not
#     OR'd word-by-word. "Virginia Tech" (as one element) returns
#     *zero* hits -- it never appears verbatim in any affiliation
#     string (they record the full official name instead) -- while
#     "Virginia" alone correctly catches that same real paper. Prefer a
#     single short, distinctive word per institution over its full
#     official/colloquial name for exactly this reason.
affiliation <- c("Colorado", "Michigan", "Rutgers", "Virginia")

candidate2 <- "Last FM" # substitute a second real candidate of your
                        # own for the batch step (10) -- a generic
                        # name like "Doe C" isn't a reliable "clean"
                        # comparison since it can collide with real,
                        # unrelated PubMed authors of that name

# All four have co-published with `candidate`; Krishnan A and Waters C
# additionally share NIH funding with `candidate` (expected in step 7).
authors <- c(
  "Brenner E",  # Evan Brenner      -- co-authorship
  "Ghosh A",    # Abhirupa Ghosh    -- co-authorship
  "Krishnan A", # Arjun Krishnan    -- co-authorship + shared NIH funding
  "Waters C"    # Christopher Waters -- co-authorship + shared NIH funding
)
```

## 1. Raw PubMed search

The lowest level: confirms the PubMed connection works and returns
PMIDs for a known name.

```{r step1-search}
ids <- pmSearchAuthor(candidate, affiliation = affiliation)
ids
```

## 2. Fetch metadata for those PMIDs

Turns PMIDs into a tidy per-(PMID, author) table -- this is what
author-name matching happens against downstream, so it's worth
eyeballing directly.

```{r step2-fetch}
pubs <- pmFetchAuthors(ids[seq_len(min(10, length(ids)))])
pubs
```

## 3. `pmCoauthors()`: steps 1+2 in one call

This is the convenience wrapper everything downstream actually uses.

```{r step3-coauthors}
pubs <- pmCoauthors(candidate, affiliation = affiliation)
pubs
```

## 4. Build the co-authorship edge list

Pairs up co-authors sharing a PMID. `candidate` has published directly
with all four of `authors`, so each should show up on an edge with
`candidate` here -- but note `edges$from`/`edges$to` are always in
reduced "Last F" form (that's what `buildCoauthorEdges()` builds them
as from PubMed's own `author_last`/`author_fore` fields), so comparing
against `candidate` directly only works because `candidate` is already
in that form for this example. A full-forename `candidate` like
`"Ravi Janani"` needs the same reduction applied before comparing --
`coiCheckR:::.pm_canonical_name()` is the (internal, `:::`-only)
function that does it, and is exactly what `checkCoi()` uses internally
so you don't have to do this by hand in normal use.

The comparison also needs to tolerate a middle initial PubMed records
on one paper but not another for the same person (e.g. "Waters C" on
one record, "Waters CM" on the next) -- `coiCheckR:::.pm_name_matches_any()`
is the prefix-tolerant matcher `checkCoi()` itself uses internally for
exactly this; a plain `==`/`%in%` here would silently under-count
relative to what `checkCoi()` actually finds in step 9.

```{r step4-edges}
edges <- buildCoauthorEdges(pubs)
candidate_canonical <- coiCheckR:::.pm_canonical_name(candidate)
candidate_canonical # "Ravi J" -- this is what actually gets compared

cand_edges <- edges |>
  dplyr::filter(
    coiCheckR:::.pm_name_matches_any(from, candidate_canonical) |
      coiCheckR:::.pm_name_matches_any(to, candidate_canonical)
  )
cand_edges

# Narrowed to just the edges connecting `candidate` to `authors`:
cand_edges |>
  dplyr::filter(
    (coiCheckR:::.pm_name_matches_any(from, candidate_canonical) &
      coiCheckR:::.pm_name_matches_any(to, authors)) |
      (coiCheckR:::.pm_name_matches_any(to, candidate_canonical) &
        coiCheckR:::.pm_name_matches_any(from, authors))
  )
```

## 5. Second-degree conflicts

Candidate hasn't published directly with one of `authors`, but shares
a frequent collaborator who has. Slower than direct co-authorship --
one PubMed query per author checked.

```{r step5-second-degree}
secondDegreeConflicts(candidate, authors, affiliation = affiliation, min_year = 2020)
```

## 6. NIH RePORTER: one PI's awards

Independent of PubMed -- confirms the funding-side lookup works on its
own. `candidate`, Krishnan A, and Waters C should each return awards
individually before step 7 cross-references them.

```{r step6-reporter}
this_year <- as.integer(format(Sys.Date(), "%Y"))
last_10y <- (this_year - 9):this_year

reporterSearchPI("Ravi, Janani", fiscal_years = last_10y)
reporterSearchPI("Krishnan, Arjun", fiscal_years = last_10y)
reporterSearchPI("Waters, Christopher", fiscal_years = last_10y)
```

## 7. NIH RePORTER: shared awards

Cross-references `candidate`'s project numbers against all four
`authors`. Expect hits against Krishnan A and/or Waters C here, even
though all four show up as co-authors in step 4 -- funding overlap and
publication overlap are independent evidence types.

```{r step7-shared-awards}
reporterSharedAwards(candidate, authors, fiscal_years = last_10y)
```

## 8. bioRxiv preprint enrichment (optional)

Only useful if you already have a DOI in hand (e.g. from a
manuscript's bibliography) -- bioRxiv has no name-search endpoint of
its own, so this one isn't keyed off `candidate`/`authors` like the
others. Substitute a real preprint DOI (e.g. one co-authored by
`candidate` and one of `authors`) to see it resolved.

```{r step8-biorxiv}
biorxivLookup("10.1101/2020.05.17.095000")
```

## 9. `checkCoi()`: everything above, combined

This is what an editor actually calls. `report$direct` should have a
row for each of the four `authors` (from step 4), and `report$funding`
should have a row for Krishnan A and/or Waters C (from step 7).

```{r step9-checkcoi}
report <- checkCoi(
  candidate, authors, affiliation = affiliation,
  check_second_degree = TRUE, check_funding = TRUE
)
report      # print.coiReport(): one-line summary + status
summary(report)
```

## 10. `checkCoiBatch()`: several candidates at once

```{r step10-batch}
batch <- checkCoiBatch(c(candidate, candidate2), authors)
batch$summary
cat(formatCoiSummary(batch), sep = "\n")
```

## 11. Visualize

```{r step11-plot, fig.width = 6, fig.height = 6}
g <- coiToGraph(report)
plotCoiNetwork(report)
```

## 12. Failure handling

The behavior this package is most careful about: disconnect from the
network, then re-run step 9. `status` should read
`"insufficient_evidence"` and `sources_failed` should name which
source(s) errored -- never silently `"no_conflict_detected"`. (The
same behavior is also covered without a live network in
`tests/testthat/test-status.R`, which mocks a `rentrez` failure via
`testthat::local_mocked_bindings()`.)

```{r step12-failure, eval = FALSE}
# eval = FALSE always, even when COICHECKR_RUN_LIVE_TESTS is set --
# running this with a live, working connection would just succeed
# normally and not demonstrate the failure path at all. Actually
# exercising this requires deliberately breaking connectivity first
# (or see tests/testthat/test-status.R for the mocked version, which
# does run automatically as part of the package's test suite).
offline_report <- checkCoi(candidate, authors)
offline_report$status          # "insufficient_evidence"
offline_report$sources_failed  # e.g. c("pubmed", "nih_reporter")
```

## What each stage tells you

Steps 1-8 are the raw evidence sources -- if one of these misbehaves,
the problem is isolated to a single API wrapper. Step 9 (`checkCoi()`)
should just be steps 4, 5, and 7 stitched together, so anything that
disagrees between step 9 and the earlier steps points at the stitching
logic in `coi_report.R` rather than the API wrappers themselves. Steps
10-11 are pure presentation over an already-built report, so bugs
there are display/formatting issues, not evidence-gathering ones.

With the example names above, expect:

- `Ravi J` x `Brenner E`, `Ghosh A`, `Krishnan A`, `Waters C` -- a
  **direct co-authorship** hit for all four (steps 4 and 9).
- `Ravi J` x `Krishnan A` and `Ravi J` x `Waters C` -- **also** a
  shared NIH funding hit (steps 6, 7, and 9), on top of the
  co-authorship above. Co-authorship and funding overlap are
  independent evidence types, so seeing both for the same pair here is
  expected, not redundant.
- Step 10's second candidate (`candidate2`) has no fixed expectation
  here -- substitute someone whose actual relationship to `authors`
  you know, and check the batch summary against that, the same way
  you're checking `candidate` above.

If any of the above doesn't match, that's the concrete signal of
where to look: a missing co-authorship hit points at steps 1-4, a
missing funding hit points at steps 6-7, and a mismatch only in step 9
(but not in the earlier steps) points at `checkCoi()`'s own stitching
logic in `coi_report.R`.

## Session info

```{r session-info}
sessionInfo()
```
